Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-09-25 manuscript/manuscript.tex
ms updates

Additionally, we looked for a signal of temporal autocovariance in
\textcite{Bergland2014-ij}, a study that collected \emph{Drosophila
melanogaster} through Spring-Fall season pairs across three years. If there was
a strong  pattern of genome-wide fluctuating selection, we might expect a
pattern of positive covariances between similar seasonal changes, e.g.
Spring-Fall in two adjacent years, and negative covariances between dissimilar
seasonal changes, e.g. Spring-Fall and Fall-Spring in two adjacent years.
However, we find no such signal over years; we discuss this in more depth in
Supplementary Materials Section \ref{bergland-reanalysis}.

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances.  To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows (Figure
\ref{fig:figure-3}, pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance of each tile will be noisy, due to sampling and genetic drift, and
the neutral distribution of the covariance is complicated due to linkage
disequilibria (which can occur over long physical distances in E\&R and
selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To address
this, we have developed a permutation-based procedure that constructs a null
distribution by randomly flipping the signs of the  allele frequency changes
per-genomic window. This destroys the systematic covariances created by linked
selection and creates a sampling distribution of the covariances spuriously
created by neutral genetic drift while preserving the complex dependencies
between adjacent loci created by linkage disequilibrium (see Supplementary
Material Section \ref{supp:empirical-null} for more details on this approach).
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

Following \textcite{Waples1989-sj}, we have that that the variance in allele
frequency change at a locus in the initial generation, which is entirely due to
the binomial sampling process, is $\var(p_0) = \nicefrac{p_0(1-p_0)}{d_0}$
where $d_0$ is the number of binomial draws (e.g. read depth). At a later
timepoint, the variance in allele frequency is a result of both the binomial
sampling process at time $t$ and the evolutionary process. Using the law of
total variation we can partition the variation from each process,

@@

\begin{align}
  V(p_t) &= V\left(p_0 + (p_1 - p_0) + (p_2 - p_1) + \ldots + (p_t - p_{t-1}) \right) \\
         &= V\left(p_0 + \Delta p_0 + \Delta p_1 + \ldots + \Delta p_{t-1} \right) \\
         &= V(p_0) + \sum_{i=0}^{t-1} \cov(p_0, \Delta p_i) + \sum_{i=0}^{t-1} \var(\Delta p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}
%

Each allele frequency change is equally like to be positive as it is to be
negative; thus by symmetry this second term is zero. Additionally $V(p_0) = 0$,
as we treat $p_0$ as a fixed initial frequency. We can write,

\begin{align}
  V(p_t) &= \sum_{i=0}^{t-1} \var(\Delta p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}

@@

\begin{align}
  V(p_t) &= \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) + \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}

@@

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= \var(\widetilde{p}_t) + \var(\widetilde{p}_0) - 2 C \cov(\widetilde{p}_t, \widetilde{p}_0) \\
                                          &= \frac{p_t(1-p_t)}{d_t}  + \frac{p_0(1-p_0)}{d_0} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right] + \\ & \;\;\;\;\;\;
                                               \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) - \frac{C p_0(1-p_0)}{2N} \\
  \frac{\var(\widetilde{p}_t - \widetilde{p}_0)}{p_0(1-p_0)} &= 1 + \frac{p_t(1-p_t)}{p_0(1-p_0)d_t}  + \frac{1}{d_0} - \left(1-\frac{1}{2N}\right)^t + \\ & \;\;\;\;\;\;
  \sum_{i=0}^{t-1} \frac{\var(\Delta_{_H} p_i)}{p_0(1-p_0)}  + \sum_{0 \le i < j}^{t-1} \frac{\cov(\Delta p_i, \Delta p_j)}{p_0(1-p_0)} - \frac{C}{N}
\end{align}
%
where $C = 1$ if Plan I is used, and $C=0$ if Plan II is used (see
\cite{Waples1989-sj}, p. 380 and Figure 1 for a description of these sampling
procedures; throughout the paper we use sampling Plan II). We move terms
creating a bias-corrected estimator for the population variance in allele
frequency change, and replace all population heterozygosity terms with the
unbiased sample estimators, e.g. $\frac{d_t}{d_t-1} \widetilde{p}_t (1-
\widetilde{p}_t)$,

@@

@@

\subsection{\textcite{Barghi2019-qy} Empirical Null and Windowed Covariance Distributions}

@@

\subsection{\textcite{Barghi2019-qy} Tail Probabilities for Windowed Covariances Distributions}

@@

\subsection{\textcite{Bergland2014-ij} Re-Analysis}
\label{bergland-reanalysis}

@@

  \caption{\textbf{A}: Histogram of original \textcite{Bergland2014-ij}
    seasonal p-values and p-values creating by randomly permuting the seasons
    at each locus. \textbf{B}: Histogram of original \textcite{Bergland2014-ij}
    p-values alongside all unique permutations (ignoring symmetries that lead
  to identical p-values).}

>>>>

Additionally, we looked for a signal of temporal autocovariance in
\textcite{Bergland2014-ij}, a study that collected \emph{Drosophila
melanogaster} through Spring-Fall season pairs across three years. If there was
a strong  pattern of genome-wide fluctuating selection, we might expect a
pattern of positive covariances between similar seasonal changes, e.g.
Spring-Fall in two adjacent years, and negative covariances between dissimilar
seasonal changes, e.g. Spring-Fall and Fall-Spring in two adjacent years.
However, we find no such signal over years; we discuss this in more depth in
Supplementary Materials Section \ref{supp:bergland-reanalysis}.

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows \vb{which
we refer to as \emph{windowed covariances}} (Figure \ref{fig:figure-3}
\vb{shows these distributions} pooling across all replicates; see Supplementary
Figure \ref{suppfig:barghi-offset-replicate-panels} for individuals
replicates). The covariance of each tile will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria (which can occur over long physical distances in
E\&R and selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To
address this, we have developed a permutation-based procedure that constructs a
null distribution by randomly flipping the signs of the  allele frequency
changes per-genomic window. This destroys the systematic covariances created by
linked selection and creates a sampling distribution of the covariances
spuriously created by neutral genetic drift while preserving the complex
dependencies between adjacent loci created by linkage disequilibrium (see
Supplementary Material Section \ref{supp:empirical-null} for more details on
this approach).  This empirical neutral null distribution is conservative in
the sense that the variances of the covariances are wider than expected under
drift alone as they include the effect of selection on the allele frequency
change within a time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

Following \textcite{Waples1989-sj}, we have that the variance in allele
frequency change at a locus in the initial generation, which is entirely due to
the binomial sampling process, is $\var(p_0) = \nicefrac{p_0(1-p_0)}{d_0}$
where $d_0$ is the number of binomial draws (e.g. read depth). At a later
timepoint, the variance in allele frequency is a result of both the binomial
sampling process at time $t$ and the evolutionary process. Using the law of
total variation we can partition the variation from each process,

@@

\begin{align}
  \var(p_t) &= \var\left(p_0 + (p_1 - p_0) + (p_2 - p_1) + \ldots + (p_t - p_{t-1}) \right) \\
         &= \var\left(p_0 + \Delta p_0 + \Delta p_1 + \ldots + \Delta p_{t-1} \right) \\
         &= \var(p_0) + \sum_{i=0}^{t-1} \cov(p_0, \Delta p_i) + \sum_{i=0}^{t-1} \var(\Delta p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}
%

Each allele frequency change is equally like to be positive as it is to be
negative; thus by symmetry this second term is zero. Additionally $\var(p_0) = 0$,
as we treat $p_0$ as a fixed initial frequency. We can write,

\begin{align}
  \var(p_t) &= \sum_{i=0}^{t-1} \var(\Delta p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}

@@

\begin{align}
  \var(p_t) &= \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) + \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i) + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).
\end{align}

@@

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= \var(\widetilde{p}_t) + \var(\widetilde{p}_0) - 2 C \cov(\widetilde{p}_t, \widetilde{p}_0) \\
                                          &= \frac{p_t(1-p_t)}{d_t}  + \frac{p_0(1-p_0)}{d_0} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right] + \\ & \;\;\;\;\;\;
                                               \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) - \frac{C p_0(1-p_0)}{2N} \\
  \frac{\var(\widetilde{p}_t - \widetilde{p}_0)}{p_0(1-p_0)} &= 1 + \frac{p_t(1-p_t)}{p_0(1-p_0)d_t}  + \frac{1}{d_0} - \left(1-\frac{1}{2N}\right)^t + \\ & \;\;\;\;\;\;
  \sum_{i=0}^{t-1} \frac{\var(\Delta_{_H} p_i)}{p_0(1-p_0)}  + \sum_{0 \le i < j}^{t-1} \frac{\cov(\Delta p_i, \Delta p_j)}{p_0(1-p_0)} - \frac{C}{N}
\end{align}
%
where $C = 1$ if Plan I is used, and $C=0$ if Plan II is used (see
\cite{Waples1989-sj}, p. 380 and Figure 1 for a description of these sampling
procedures; throughout the paper we use sampling Plan II). Rearranging, we can
create a bias-corrected estimator for the population variance in allele
frequency change, and replace all population heterozygosity terms with the
unbiased sample estimators, e.g. $\frac{d_t}{d_t-1} \widetilde{p}_t (1-
\widetilde{p}_t)$,

@@

XXX

@@

\clearpage
\subsection{\textcite{Barghi2019-qy} Empirical Null and Windowed Covariance Distributions}

@@

\clearpage
\subsection{\textcite{Barghi2019-qy} Tail Probabilities for Windowed Covariances Distributions}

@@

\subsection{\textcite{Bergland2014-ij} Re-Analysis}
\label{supp:bergland-reanalysis}

@@

  \caption{\vb{\textbf{A}: Histogram of original \textcite{Bergland2014-ij}
    seasonal p-values and p-values creating by randomly permuting the seasons
    at each locus. \textbf{B}: Histogram of original \textcite{Bergland2014-ij}
    p-values alongside all unique permutations (ignoring symmetries that lead
    to identical p-values).}}
